---
title: "Linear Mixed Model 1"
author: "Amy Kim - atykim@ucdavis.edu"
date: "10/30/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Linear Mixed Model
$$\mathbf{Y} = \mathbf{X\beta} + \mathbf{Z\alpha} + \mathbf{\epsilon}$$
where $E(\alpha_i) = 0, Var(\alpha_i) = \sigma^2, E(\epsilon_{ij}) = 0, Var(\epsilon_{ij}) = \tau^2$ 

Comparing with Linear Model $\mathbf{Y} = \mathbf{X\beta} + \mathbf{\epsilon}$

##0. Estimator

###(a). Maximum Likelihood Estimator (MLE)

###(b). Restricted Maximum Likelihood Estimator (REML)

##1. Example

```{r, echo = FALSE}
library(nlme)
library(lme4)
library(knitr)
#Read the data
corps = read.table("./Data/corps.txt", header = TRUE)
chairman = factor(rep(1:50, each = 2)) #factor: random effect
corps$chairman = chairman
corps$year = as.factor(corps$year)
```

#REML

##a.

```{r, warning=FALSE}
setwd("~/OneDrive/STA232B16")
library(knitr)
#setting up data
corps = read.table("corps.txt", header = TRUE)
chairman = factor(rep(1:50, each = 2))
corps$chairman = chairman
corps$year = as.factor(corps$year)

#reml
mod1 = nlme::lme(fixed = y ~ year + AGE + SHARES + REV + INC, random = ~ 1|chairman,
                 data = corps, method = "REML")
mod1l = lme4::lmer(formula = y ~ year + AGE + SHARES + REV + INC + (1|chairman),
                   data = corps, REML = TRUE)
summary(mod1)

#estimators 
sigma2 = HLMdiag::varcomp.mer(mod1l)[2] #nlme::getVarCov(mod1)
tau2 = HLMdiag::varcomp.mer(mod1l)[1] # summary(mod1)$sigma^2
varcomp = c(sigma2, tau2)
names(varcomp) = c("sigma2", "tau2")
kable(varcomp) #estimators for tau^2 and sigma^2 (according to project notation)
nlme::fixef(mod1) #fixed estimators
```

##b. 
Explain how the standard error obtained for the REML estimates of the fixed effects.
Write some simple R code yourself to verify the standard error results.

```{r}
X = as.matrix(lme4::getME(mod1l, 'X'))
Z = as.matrix(lme4::getME(mod1l, 'Z'))
V = sigma2*Z %*% t(Z) + tau2 * diag(100)
v = solve(V)
covmatb = solve(t(X) %*% v %*% X)

#Covariance Matrix
kable(covmatb) #by matrix calculation
kable(vcov(mod1)) #covariance matrix by R

sqrt(diag(covmatb))
sqrt(diag(vcov(mod1)))
```

##c.
Explain two methods for obtaining the standard errors for the variance-component.

 * Asymptotic Covariance Matrix
    + We can find inverse of fisher information matrix which is aysmptotically equivalent to covariance Matrix.
 * Bootstrapping Method
    + We can use parameter bootstrap methods to simulate get many estimators. Then, we calculate standard error of bootstrapped estimators.
 
##d.
Compute Standard Error by ACM methods.
```{r}
#first find the fisher's information matrix
P = v - v %*% X %*% solve(t(X) %*% v %*% X) %*% t(X) %*% v
I = matrix(1:4, 2, 2)
tmpZ = Z %*% t(Z)
I[1,1] = 0.5 * sum(diag(P %*% tmpZ %*% P %*% tmpZ))
I[1,2] = I[2,1] = 0.5 * sum(diag(P %*% P %*% tmpZ))
I[2,2] = 0.5 * sum(diag(P %*% P))

#get ACM
acm = solve(I)
seVCa = sqrt(diag(acm))
names(seVCa) = c("se(sigma)", "se(tau)")
kable(seVCa)
```

##e.
Compute Standard Error by Bootstrapping

```{r, warning=FALSE}
beta = nlme::fixef(mod1)
#sigma2 ,tau2
set.seed(12043)
newY = MASS::mvrnorm(100, X%*%beta, V) #each row has one sample, the number of row = sample size

vcM = apply(newY, 1, function(x){
  fit = lme4::lmer(formula = x ~ year + AGE + SHARES + REV + INC + (1|chairman),
                   data = corps, REML = TRUE);
  HLMdiag::varcomp.mer(fit)
})
tmp1 = apply(vcM, 1, mean)
tmp2 = apply(vcM, 1, sd)
#comparing to ACM
seVCb = c(tmp2[2], tmp2[1])
names(seVCb) = c("se(sigma)", "se(tau)")
seVC_RM = rbind(seVCb, seVCa)
row.names(seVC_RM) = c('boot', 'ACM')
kable(seVC_RM)

#beta
betaM = apply(newY, 1, function(x){
  fit = lme4::lmer(formula = x ~ year + AGE + SHARES + REV + INC + (1|chairman),
                   data = corps, REML = TRUE);
  lme4::fixef(fit);
})
apply(betaM, 1, mean)
apply(betaM, 1, sd)

```

#ML

##a. Obtain ML estimators

```{r, warning=FALSE}
#reml
mod2 = nlme::lme(fixed = y ~ year + AGE + SHARES + REV + INC, random = ~ 1|chairman,
                 data = corps, method = "ML")
mod2l = lme4::lmer(formula = y ~ year + AGE + SHARES + REV + INC + (1|chairman),
                   data = corps, REML = FALSE)
#nlme::fixef(mod2)
summary(mod2)

#estimators 
sigma2m = HLMdiag::varcomp.mer(mod2l)[2] #nlme::getVarCov(mod2)
tau2m = HLMdiag::varcomp.mer(mod2l)[1] # summary(mod2)$sigma^2
varcomp.m = c(sigma2m, tau2m)
names(varcomp.m) = c("sigma2", "tau2")
kable(varcomp.m) #estimators for tau^2 and sigma^2 (according to project notation)
kable(varcomp)
nlme::fixef(mod2) #fixed estimators
```
The ML fixed effect estimators are same as the REML fixed effect estimators; but the estimators of variance componants by ML and that by REML are different. (ML one is smaller)

##d.
Compute Standard Error by ACM methods.
```{r}
#first find the fisher's information matrix
Vm = sigma2m*Z %*% t(Z) + tau2m * diag(100)
vm = solve(Vm)
Im = matrix(1:4, 2, 2)
Im[1,1] = 0.5 * sum(diag(vm %*% tmpZ %*% vm %*% tmpZ))
Im[1,2] = Im[2,1] = 0.5 * sum(diag(vm %*% vm %*% tmpZ))
Im[2,2] = 0.5 * sum(diag(vm %*% vm))

#get ACM
acm.m = solve(Im)
seVCa.m = sqrt(diag(acm.m))
names(seVCa.m) = c("se(sigma)", "se(tau)")
kable(seVCa.m)
```

##e.
Compute Standard Error by Bootstrapping

```{r, warning=FALSE}
beta2 = nlme::fixef(mod2)
#sigma2 ,tau2
set.seed(12043)
newY = MASS::mvrnorm(100, X%*%beta2, Vm) #each row has one sample, the number of row = sample size

vcM.m = apply(newY, 1, function(x){
  fit = lme4::lmer(formula = x ~ year + AGE + SHARES + REV + INC + (1|chairman),
                   data = corps, REML = FALSE);
  HLMdiag::varcomp.mer(fit)
})
tmp1.m = apply(vcM.m, 1, mean)
tmp2.m = apply(vcM.m, 1, sd)

#comparing to ACM
seVCb.m = c(tmp2.m[2], tmp2.m[1])
names(seVCb.m) = c("se(sigma)", "se(tau)")
seVC_M = rbind(seVCb.m, seVCa.m)
row.names(seVC_M) = c('boot', 'ACM')
kable(seVC_M)

#beta
betaM.m = apply(newY, 1, function(x){
  fit = lme4::lmer(formula = x ~ year + AGE + SHARES + REV + INC + (1|chairman),
                   data = corps, REML = TRUE);
  lme4::fixef(fit);
})
apply(betaM.m, 1, mean)
apply(betaM.m, 1, sd)
```



## Appendix

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
